---
title: "GP Theory Part 2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(mnormt)
library(plgp)
library(reshape2)
set.seed(02092021)
```


#### Correlation function: more details

Recall the variant of the exponential covariance function that we have previously seen. Where $d$ as the Euclidean distance between $x_1$ and $x_2$, such that $d = \sqrt{(x_i - x_j)^2}$

$$\rho_{i,j} = \exp \left(- d \right)$$

\vfill

Lets view the exponential correlation as a function of distance between the two points.

```{r, echo = F}
dist <- seq(0, 5, by = .1)

tibble(rho = exp(-dist), dist = dist) %>%
  ggplot(aes(y = rho, x = dist)) + geom_line() +
  theme_bw() + ylab(expression(rho))

```

\vfill


Now let's consider a more general framework for covariance where 

$$\sigma_{i,j} = \sigma^2 \exp \left(- d_{ij} /\phi \right)$$
\newpage

Now we have introduced two new parameters into this function. What do you suppose that they do?

- $\sigma^2$: controls the magnitude of the covariance.

\vfill

- $\phi$: controls the range of the spatial correlation

\vfill

Modify the following code (from last lecture) to gain an intuition about these parameters.
\vfill

```{r}
phi <- 1
sigmasq <- 1
x <- seq(0, 10, by = .1)
n <- length(x)
d <- plgp::distance(x)
eps <- sqrt(.Machine$double.eps) 
H <- exp(-d/phi) + diag(eps, n) 
H[1:3,1:3]
y <- rmnorm(1, rep(0,n),sigmasq * H)
tibble(y = y, x = x) %>% ggplot(aes(y=y, x=x)) +
  geom_line() + theme_bw() + ggtitle('Random realization of a GP with phi = 1 and sigmasq = 1') +
  geom_point(size = .5)
```

\newpage

```{r}
phi <- .1
sigmasq <- 5
H <- exp(-d/phi) + diag(eps, n) 
H[1:3,1:3]
y <- rmnorm(1, rep(0,n),sigmasq * H)
tibble(y = y, x = x) %>% ggplot(aes(y=y, x=x)) +
  geom_line() + theme_bw() + ggtitle('Random realization of a GP with phi = .1 and sigmasq = 5') +
  geom_point(size = .5)
```

\newpage

We will soon talk about a more broad set of correlation functions and another parameter that provides flexibility so that predictions do not have to directly through observed points.

\vfill

### Geostatistical Data
At last, we will look at simulated 2-d "spatial" data.


##### 1. Sampling Locations

```{r}
num.locations <- 50
coords <- data.frame(x = runif(num.locations), y = runif(num.locations))
coords %>% ggplot(aes(x=x,y=y)) + geom_point() + 
  ggtitle('Hypothetical Sampling Locations') + xlim(0,1) + 
  ylim(0,1) + theme_bw()
```

##### 2. Calculate Distances

```{r, echo = T}
dist.mat <- plgp::distance(coords)
```

##### 3. Define Covariance Function and Set Parameters
Use exponential covariance with no nugget: 

```{r}
sigma.sq <- 1
phi <- .1
Sigma <- sigma.sq * exp(- dist.mat/phi) + diag(eps, num.locations) 
```


##### 4. Sample realization of the process
- This requires a distributional assumption, we will use the Gaussian distribution

```{r}
Y <- rmnorm(n=1, mean = 0, varcov = Sigma)
```

- What about the rest of the locations on the map?

##### 5. Vizualize Spatial Process

```{r}
coords %>% mutate(Y = Y) %>% ggplot(aes(x=x,y=y)) + geom_point(aes(color=Y), size=2) + 
  ggtitle(label = 'Simulated Spatial Process', 
          subtitle = 'Exponential Covariance: sigma.sq = 1, phi = .1') +  
  xlim(0,1) + ylim(0,1) +   scale_colour_gradient2() + theme_dark()
```

\newpage

Now we can look at more sampling locations

```{r, echo = F}
dim.grid <- 10
grid.coords <- data.frame(x.grid = rep(seq(.05, .95, length.out=dim.grid), dim.grid),
  y.grid = rep(seq(.05, .95, length.out = dim.grid), each = dim.grid))

dist.grid <- distance(grid.coords)

sigma.sq <- 1
phi <- .1
Sigma <- sigma.sq * exp(- dist.grid/phi) + diag(eps, dim.grid ^ 2) 

Y <- rmnorm(n=1, mean = 0, varcov = Sigma)

grid.coords %>% mutate(Y = Y) %>% ggplot(aes(x=x.grid,y=y.grid)) + geom_point(aes(color=Y), size=3) + ggtitle('Simulated Spatial Process', subtitle = 'Exponential Covariance: sigma.sq = 1, phi = .1') + xlim(0,1) + ylim(0,1) +   scale_colour_gradient2() + theme_dark()

```


```{r, echo = F}
dim.grid <- 50
grid.coords <- data.frame(x.grid = rep(seq(.05, .95, length.out=dim.grid), dim.grid),
  y.grid = rep(seq(.05, .95, length.out = dim.grid), each = dim.grid))

dist.grid <- distance(grid.coords)

sigma.sq <- 1
phi <- .1
Sigma <- sigma.sq * exp(- dist.grid/phi) + diag(eps, dim.grid ^ 2) 

Y <- rmnorm(n=1, mean = 0, varcov = Sigma)

grid.coords %>% mutate(Y = Y) %>% ggplot(aes(x=x.grid,y=y.grid)) + geom_point(aes(color=Y), size=3) + ggtitle('Simulated Spatial Process', subtitle = 'Exponential Covariance: sigma.sq = 1, phi = .1') + xlim(0,1) + ylim(0,1) +   scale_colour_gradient2() + theme_dark()
```




How does the spatial process change with:

- another draw with same parameters?
- a different value of $\phi$
- a different value of $\sigma^2$

